A user-extensible and adaptable parser architecture

### 2008-08-15 doc/sgai-2008.tex
Finished my corrections.

There is some similarity between the architecture and William Wood's
Augmented Transition Networks (ATN)~\cite{atns, nlpip}, a tool used in
Computational Linguistics for creating grammars to parse or generate
sentences.  The resemblance between ATN and the parser is accidental, but
it is interesting how two different approaches have a similar division of
responsibilities, while having different semantics.

@@

\begin{table}[ht]
    \caption{Similarities with ATN}\label{Similarities with ATN}
    % The \smallskip{} below stop the text hitting the lines.
    \begin{tabular}[]{lll}
        \hline
        \noalign{\smallskip}
        ATN           & Parser    & Similarity                          \\
        \noalign{\smallskip}
        \hline
        \noalign{\smallskip}
        Networks      & Framework & Determines the sequence of
                                    transitions or actions which        \\
                      &           & constitutes a valid input.          \\
        Transitions   & Actions   & Save data and impose conditions the
                                    input must meet to be               \\
                      &           & considered valid.                   \\
        Abbreviations & Rules     & Responsible for classifying input.  \\
        \noalign{\smallskip}
        \hline
        \noalign{\smallskip}
    \end{tabular}
\end{table}

@@

Each action contains the code required to deal with a particular type of
log lines, e.g.\ rejections.  The actions are parser-specific: each parser
author will need to write the actions required from scratch, unless
extending an existing parser.  In the Postfix parser developed for this
project there are \numberOFactions{} actions and \numberOFrules{} rules,
with a very skewed distribution of rules to actions as shown in
\graphref{Distribution of rules per action}.  The distribution is skewed
for the same reason the architecture is so successful in enabling users to
parse new log lines: most new log lines can be parsed by a combination of a
new rule and an existing action.  It is expected that parsers utilising
this architecture will have a high ratio of rules to actions, hopefully
resulting in simpler rules and clearer distinctions between log lines.
\showgraph{build/graph-action-distribution}{Distribution of rules per
action}{Distribution of rules per action} Unsurprisingly the action with
the highest number of associated rules is \texttt{REJECTION}, the action
which handles Postfix rejecting a mail; it is followed by
\texttt{SAVE\_BY\_QUEUEID}, the action responsible for handling informative
inputs: these are log lines that supplement the data gathered from other
log lines (such as rejections), but that do not by themselves represent
interesting events.  The third most common action is, perhaps surprisingly,
\texttt{IGNORE}: the \texttt{IGNORE} action does nothing when executed,
allowing uninteresting log lines to be parsed without causing any effects
(neither \texttt{REJECTION} nor \texttt{IGNORE} imply that the input is
ungrammatical or unparsed).  The parser warns about every unparsed log
line, to alert the user that they need to alter or extend their ruleset;
the Postfix parser successfully parses every input line in the
\numberOFlogFILESall{} log files it is currently being tested with.  The
remaining actions have only one or two associated rules: in some cases this
is because there will only ever be one type of log line for that action,
e.g.\ all log lines showing that a remote client has connected are matched
by a single rule and handled by the \texttt{CONNECT} action; in other cases
the action is required to address a deficiency in the log files, or a
complication which arises during parsing.  Using the \texttt{CONNECT}
action as an example, it creates a new data structure for each client
connecting, and saves the data extracted by the rule.  If the data
structure already exists it is treated as a symptom of a bug, and the
action issues a warning containing the full contents of the data structure,
plus the log line which has just been parsed.

@@

During development of the Postfix parser it became apparent that in
addition to the obvious variety in log lines, there were a large number of
complications to be overcome.  Some were the result of deficiencies in
Postfix's logging (some of which were improved on or resolved by later
versions of Postfix); others were due to the vagaries of process
scheduling, client behaviour, and administrative actions.  All were
successfully accommodated in the Postfix parser: adding new actions was
enough to overcome several of the complications; others required
modifications to a single existing action to work around the difficulties;
the remainder were resolved by adapting actions to pass information to
subsequent actions (via the framework), which check for the extra
information and change their behaviour appropriately.

@@

Rules categorise inputs, specifying both the regex to match against each
log line and the action to invoke when the match is successful.  The
Postfix parser stores its rules in the same SQL database the results are
stored in, removing any doubt as to which set of rules was used to produce
a set of results; other implementations are free to store the rules in
whatever fashion suits their needs.  The framework requires each rule to
have \texttt{action} and \texttt{regex} attributes; each implementation is
free to add any additional attributes it requires.  The Postfix parser adds
several attributes to optimise rule ordering (\sectionref{Rule ordering}),
to restrict which log lines rules can be matched against, and to describe
each rule.  Data is automatically extracted and saved from the log line
based on keywords in the rule's regex, but there is also a facility to
specify extra data to be saved.

Adding new rules is simple: define all the required attributes and append
the new rule to wherever the existing rules are stored.  The Postfix parser
supplies a utility based on SLCT~\cite{slct-paper} to aid in producing
regexes from previously unparsed inputs.  Storing the rules separately
allows other approaches to rule management to be used, e.g.\ instead of
manually adding new rules when unparsed log lines are seen, machine
learning techniques could be used to automatically generate new rules,
based on the existing rules and input lines.  If this approach was taken
the choice of machine learning technique would be constrained by the size
of typical data sets: the \numberOFlogFILES{} log files used to generate
statistics contain \numberOFlogLINEShuman{} log lines.  Techniques which
require the full data set when training would be impractical; Instance
Based Learning~\cite{instance-based-learning} techniques which
automatically determine which inputs from the training set are valuable and
which inputs can be discarded would probably reduce the data required to a
manageable size.  This avenue of research and development has not been
pursued by the authors, but could easily be independently undertaken.

@@

The parser does not try to detect overlapping rules: that responsibility is
left to the author of the rules.  Unintentionally overlapping rules lead to
inconsistent parsing and data extraction because the first matching rule
wins, and the order in which rules are tried against each line may change
between log files.  Overlapping rules are frequently a requirement,
allowing a more specific rule to match some log lines and a more general
rule to match the majority, e.g.\ separating SMTP delivery to specific
sites from SMTP delivery to the rest of the world.  Overlapping rules
should have a priority attribute to specify the order in which they should
be used; negative priorities may be useful for catchall rules.

Storing the rules separately to the parser allows external tools to be
written to detect overlapping rules.  Traditional regexes are equivalent in
computational power to Finite Automata (FA); it is possible to convert
between regexes and FA, so regex overlap could be detected by finding a
non-empty intersection of two FA\@.  Perl 5.10 regexes are more powerful
than traditional regexes, e.g.\ it is possible to match correctly nested
and balanced brackets to an arbitrary depth:
\newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
Perl 5.10 regexes can maintain an arbitrary state stack and are thus
equivalent in computational power to Pushdown Automata (PA), so overlap
detection may require calculating the intersection of two PA.\@  Detecting
overlap between $n$ regexes requires $\frac{n(n-1)}{2}$ intersections to be
calculated, giving $O(n^2x)$ complexity, where $O(x)$ is the complexity of
calculating intersection.  This is certainly not a task to be performed on
every parsing run: detecting overlap amongst the Postfix parser's
\numberOFrules{} rules would require calculating
\numberOFruleINTERSECTIONS{} intersections.

@@

Parsing efficiency is an obvious concern when the parser routinely needs to
parse very large log files.  The mail server which generated the log files
used in testing this parser accepts approximately 10,000 mails for 700
users a day.  Median log file size is 50MB, containing 285,000 log lines
--- large scale mail servers would have much larger log files.  When
generating the timing data used in this section, \numberOFlogFILES{} log
files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
parsed 10 times and the parsing times averaged.  Saving results to the
database was disabled for the test runs, as that dominates the run time of
the program, and the tests are aimed at measuring the speed of the parser
rather than the speed of the database and the disks the database is stored
on.  The computer used for test runs was a Dell Optiplex 745 described in
\tableref{Computer used to generate statistics}.  Parsing all
\numberOFlogFILES{} log files in one run took
\input{build/include-timing-run-duration}.  The computer was dedicated to
the task of gathering statistics from test runs, and was not used for any
other purpose while test runs were ongoing; any services not necessary for
running the tests were disabled.

@@

An important property of a parser is how parsing time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
\Graphref{Parsing time, file size, and number of lines} shows the parsing
time in seconds, file size in MB, and number of log lines in tens of
thousands, for each of the \numberOFlogFILES{} log files.  All three lines
run roughly in parallel, giving the impression that the algorithm scales
linearly with input size.  This impression is borne out by
\graphref{parsing time vs file size vs number lines factor} which plots
both the ratio of file size vs parsing time, and the ratio of number of log
lines vs parsing time (higher is better in both cases).  The ratios are
quite tightly banded across the graph, showing that the algorithm scales
linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
despite their larger than average size (shown in \graphref{Parsing time,
file size, and number of lines}).

@@

@@

How any architecture scales as the number of rules increases is important,
but it is particularly important in this architecture because it is
expected that there will be a large number of rules.  There are
\numberOFrules{} rules in the full Postfix parser ruleset, whereas the
minimum number of rules required to parse the \numberOFlogFILES{} log files
is \numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
ruleset.  A second set of statistics was generated using the minimum
ruleset and the parsing times compared to those generated using the full
ruleset: the percentage parsing time increase when using the full ruleset
instead of the minimal ruleset for optimal, shuffled and reversed orderings
is shown in \graphref{Percentage parsing time increase of maximum ruleset
over minimum ruleset}.
\showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}

@@

The parser's coverage of Postfix log files is separated into two parts: log
line coverage and mail coverage.  The former is initially more important,
as the parser must successfully parse every line if it is to be complete,
but subsequently the latter takes precedence because reproducing the path a
mail takes through Postfix is the aim of the parser.  Improving the former
is less intrusive, as it just requires new rules to be written; improving
the latter is more intrusive as it requires changes to actions, and it can
also be much harder to notice a deficiency.

@@

Coverage of mails is much more difficult to determine accurately than
coverage of log lines.  The parser can dump its state tables in a human
readable form; examining these tables with reference to the log files is
the best way to detect mails which were not handled properly.  The parser
issues warnings when it detects any errors, some of which may alert the
user to a problem.  There should be few or no warnings when parsing, and
when finished parsing the state table should only contain entries for mails
which had yet to be delivered when the log files ended, or were accepted
before the log files began.  A sample of 6000 log lines was parsed with all
debugging options enabled, resulting in 167,448 lines of output.  All
167,448 lines were examined in conjunction with the input log file and a
dump of the resulting database, verifying that for each of the input lines
the parser used the correct rule and executed the correct action, which in
turn produced the correct result and inserted the correct data in the
database.  The log segment produced 4 warnings (about mails which had
started before the log file segment), 10 mails remaining in the state
tables, and 1625 connections correctly entered in the database.  No error
messages were produced, therefore there were no false negatives.  Given the
evidence detailed above, the authors are confident that the false positive
rate when reconstructing a mail is zero.

@@

The dataset gathered by the parser provides the foundation for the future
of this project: applying machine-learning algorithms to the data to
analyse and optimise the set of rules in use, followed by identifying
patterns in the data which could be used to write new filters to recognise
and reject spam rather than accepting it.  The parser provides the data in
a normalised form which is far easier to use as input to new or existing
implementations of machine-learning algorithms than trying to adapt each
algorithm to extract data directly from the log files.  The current focus
is on clustering and decision trees to optimise the order in which rules
are applied; future efforts will involve using data gathered by the parser
to train and test new filters.  This task is very similar to optimising a
black box application based on its inputs and outputs, and this approach
could be applied to optimising the behaviour of any system given sufficient
logging to analyse.  An alternative approach to black box optimisation
which uses application profiling in conjunction with the application's
error messages to improve the error messages shown to users is described
in~\cite{black-box-error-reporting}; profiling data may be useful in
supplementing systems which fail to provide adequate log messages.

>>>>

There is some similarity between the architecture and William Wood's
Augmented Transition Networks (ATN)~\cite{atns, nlpip}, a tool used in
Computational Linguistics for creating grammars to parse or generate
sentences.  The resemblance between ATN and the architecture is accidental,
but it is interesting how two different approaches have a similar division
of responsibilities, while having different semantics.

@@

\begin{table}[ht]
    \caption{Similarities with ATN}\label{Similarities with ATN}
    % The \smallskip{} below stop the text hitting the lines.
    \begin{tabular}[]{lll}
        \hline
        \noalign{\smallskip}
        ATN           & Architecture    & Similarity                          \\
        \noalign{\smallskip}
        \hline
        \noalign{\smallskip}
        Networks      & Framework       & Determines the sequence of
                                          transitions or actions which        \\
                      &                 & constitutes a valid input.          \\
        Transitions   & Actions         & Save data and impose conditions the
                                          input must meet to be               \\
                      &                 & considered valid.                   \\
        Abbreviations & Rules           & Responsible for classifying input.  \\
        \noalign{\smallskip}
        \hline
        \noalign{\smallskip}
    \end{tabular}
\end{table}

@@

Each action contains the code required to deal with a particular type of
log lines, e.g.\ rejections.  The actions are parser-specific: each parser
author will need to write the actions required from scratch, unless
extending an existing parser.  In the Postfix log parser developed for this
project there are \numberOFactions{} actions and \numberOFrules{} rules,
with a very skewed distribution of rules to actions as shown in
\graphref{Distribution of rules per action}.  The distribution is skewed
for the same reason the architecture is so successful in enabling users to
parse new log lines: most new log lines can be parsed by a combination of a
new rule and an existing action.  It is expected that parsers utilising
this architecture will have a high ratio of rules to actions, hopefully
resulting in simpler rules and clearer distinctions between log lines.
\showgraph{build/graph-action-distribution}{Distribution of rules per
action}{Distribution of rules per action} Unsurprisingly the action with
the highest number of associated rules is \texttt{REJECTION}, the action
which handles Postfix rejecting a mail; it is followed by
\texttt{SAVE\_BY\_QUEUEID}, the action responsible for handling informative
inputs: these are log lines that supplement the data gathered from other
log lines (such as rejections), but that do not by themselves represent
interesting events.  The third most common action is, perhaps surprisingly,
\texttt{IGNORE}: the \texttt{IGNORE} action does nothing when executed,
allowing uninteresting log lines to be parsed without causing any effects
(neither \texttt{REJECTION} nor \texttt{IGNORE} imply that the input is
ungrammatical or unparsed).  The framework warns about every unparsed log
line, to alert the user that they need to alter or extend their ruleset;
the Postfix log parser successfully parses every input line in the
\numberOFlogFILESall{} log files it is currently being tested with.  The
remaining actions have only one or two associated rules: in some cases this
is because there will only ever be one type of log line for that action,
e.g.\ all log lines showing that a remote client has connected are matched
by a single rule and handled by the \texttt{CONNECT} action; in other cases
the action is required to address a deficiency in the log files, or a
complication which arises during parsing.  Using the \texttt{CONNECT}
action as an example, it creates a new data structure for each client
connecting, and saves the data extracted by the rule.  If the data
structure already exists it is treated as a symptom of a bug, and the
action issues a warning containing the full contents of the data structure,
plus the log line which has just been parsed.

@@

During development of the Postfix log parser it became apparent that in
addition to the obvious variety in log lines, there were a large number of
complications to be overcome.  Some were the result of deficiencies in
Postfix's logging (some of which were improved on or resolved by later
versions of Postfix); others were due to the vagaries of process
scheduling, client behaviour, and administrative actions.  All were
successfully accommodated in the Postfix log parser: adding new actions was
enough to overcome several of the complications; others required
modifications to a single existing action to work around the difficulties;
the remainder were resolved by adapting actions to pass information to
subsequent actions (via the framework), which check for the extra
information and change their behaviour appropriately.

@@

Rules categorise inputs, specifying both the regex to match against each
log line and the action to invoke when the match is successful.  The
Postfix log parser stores its rules in the same SQL database the results
are stored in, removing any doubt as to which set of rules was used to
produce a set of results; other implementations are free to store the rules
in whatever fashion suits their needs.  The framework requires each rule to
have \texttt{action} and \texttt{regex} attributes; each implementation is
free to add any additional attributes it requires.  The Postfix log parser
adds several attributes to optimise rule ordering (\sectionref{Rule
ordering}), to restrict which log lines rules can be matched against, and
to describe each rule.  Data is automatically extracted and saved from the
log line based on keywords in the rule's regex, but there is also a
facility to specify extra data to be saved.

Adding new rules is simple: define all the required attributes and append
the new rule to wherever the existing rules are stored.  The Postfix log
parser supplies a utility based on SLCT~\cite{slct-paper} to aid in
producing regexes from previously unparsed inputs.  Storing the rules
separately allows other approaches to rule management to be used, e.g.\
instead of manually adding new rules when unparsed log lines are seen,
machine learning techniques could be used to automatically generate new
rules, based on the existing rules and input lines.  If this approach was
taken the choice of machine learning technique would be constrained by the
size of typical data sets: the \numberOFlogFILES{} log files used to
generate statistics contain \numberOFlogLINEShuman{} log lines.  Techniques
which require the full data set when training would be impractical;
Instance Based Learning~\cite{instance-based-learning} techniques which
automatically determine which inputs from the training set are valuable and
which inputs can be discarded would probably reduce the data required to a
manageable size.  This avenue of research and development has not been
pursued by the authors, but could easily be independently undertaken.

@@

The architecture does not try to detect overlapping rules: that
responsibility is left to the author of the rules.  Unintentionally
overlapping rules lead to inconsistent parsing and data extraction because
the first matching rule wins, and the order in which rules are tried
against each line may change between log files.  Overlapping rules are
frequently a requirement, allowing a more specific rule to match some log
lines and a more general rule to match the majority, e.g.\ separating SMTP
delivery to specific sites from SMTP delivery to the rest of the world.
This can simplify both the general rule and the more specific rule;
additionally rules from different sources can be combined with a minimum of
prior cooperation or modification required.  Overlapping rules should have
a priority attribute to specify the order in which they should be used;
negative priorities may be useful for catchall rules.

Storing the rules separately to the parser allows external tools to be
written to detect overlapping rules.  Traditional regexes are equivalent in
computational power to Finite Automata (FA); it is possible to convert
between regexes and FA, so regex overlap could be detected by finding a
non-empty intersection of two FA\@.  Perl 5.10 regexes are more powerful
than traditional regexes, e.g.\ it is possible to match correctly nested
and balanced brackets to an arbitrary depth:
\newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
Perl 5.10 regexes can maintain an arbitrary state stack and are thus
equivalent in computational power to Pushdown Automata (PA), so overlap
detection may require calculating the intersection of two PA.\@  Detecting
overlap between $n$ regexes requires $\frac{n(n-1)}{2}$ intersections to be
calculated, giving $O(n^2x)$ complexity, where $O(x)$ is the complexity of
calculating intersection.  This is certainly not a task to be performed on
every parsing run: detecting overlap amongst the Postfix log parser's
\numberOFrules{} rules would require calculating
\numberOFruleINTERSECTIONS{} intersections.

@@

Parsing efficiency is an obvious concern when the Postfix log parser
routinely needs to parse very large log files.  The mail server which
generated the log files used in testing the Postfix log parser accepts
approximately 10,000 mails for 700 users a day.  Median log file size is
50MB, containing 285,000 log lines --- large scale mail servers would have
much larger log files.  When generating the timing data used in this
section, \numberOFlogFILES{} log files (totaling 10.08 GB,
\numberOFlogLINEShuman{} log lines) were each parsed 10 times and the
parsing times averaged.  Saving results to the database was disabled for
the test runs, as that dominates the run time of the program, and the tests
are aimed at measuring the speed of the Postfix log parser rather than the
speed of the database and the disks the database is stored on.  The
computer used for test runs was a Dell Optiplex 745 described in
\tableref{Computer used to generate statistics}.  Parsing all
\numberOFlogFILES{} log files in one run took
\input{build/include-timing-run-duration}.  The computer was dedicated to
the task of gathering statistics from test runs, and was not used for any
other purpose while test runs were ongoing; any services not necessary for
running the tests were disabled.

@@

An important property of a parser is how parsing time scales relative to
input size: does it scale linearly, polynomially, or exponentially?
\Graphref{Parsing time, file size, and number of lines} shows the parsing
time in seconds, file size in MB, and number of log lines in tens of
thousands, for each of the \numberOFlogFILES{} log files.  All three lines
run roughly in parallel, giving the impression that the algorithm scales
linearly with input size.  This impression is borne out by
\graphref{parsing time vs file size vs number lines factor}: the ratios are
quite tightly banded across the graph, showing that the algorithm scales
linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
despite their larger than average size (due to a mail forwarding loop
resulting in a greatly increased number of mails delivered and log lines
generated).

@@

% XXX MAYBE ADD A TABLE SHOWING MEAN AND STDDEV OF THE RATIOS?

@@

How any architecture scales as the number of rules increases is important,
but it is particularly important in this architecture because it is
expected that there will be a large number of rules.  There are
\numberOFrules{} rules in the full Postfix log parser ruleset, whereas the
minimum number of rules required to parse the \numberOFlogFILES{} log files
is \numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
ruleset.  A second set of statistics was generated using the minimum
ruleset and the parsing times compared to those generated using the full
ruleset: the percentage parsing time increase when using the full ruleset
instead of the minimal ruleset for optimal, shuffled and reversed orderings
is shown in \graphref{Percentage parsing time increase of maximum ruleset
over minimum ruleset}.
\showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}{Percentage parsing
time increase of maximum ruleset over minimum ruleset}

@@

The Postfix log parser's coverage of log files is separated into two parts:
log line coverage and mail coverage (the former is a requirement for the
latter to be achieved).  Improving the former is less intrusive, as it just
requires new rules to be written; improving the latter is more intrusive as
it requires changes to actions, and it can also be much harder to notice a
deficiency.

@@

Coverage of mails is much more difficult to determine accurately than
coverage of log lines.  The implementation can dump its state tables in a
human readable form; examining these tables with reference to the log files
is the best way to detect mails which were not handled properly.  The
implementation issues warnings when it detects any errors, some of which
may alert the user to a problem.  There should be few or no warnings when
parsing, and when finished parsing the state table should only contain
entries for mails which had yet to be delivered when the log files ended,
or were accepted before the log files began.  A sample of 6000 log lines
was parsed with all debugging options enabled, resulting in 167,448 lines
of output.  All 167,448 lines were examined in conjunction with the input
log file and a dump of the resulting database, verifying that for each of
the input lines the implementation used the correct rule and executed the
correct action, which in turn produced the correct result and inserted the
correct data in the database.  The log segment produced 4 warnings (about
mails which had started before the log file segment), 10 mails remaining in
the state tables, and 1625 connections correctly entered in the database.
No error messages were produced, therefore there were no false negatives.
Given the evidence detailed above, the authors are confident that the false
positive rate when reconstructing a mail is zero.

@@

The dataset gathered by the Postfix log parser provides the foundation for
the future of this project: applying machine-learning algorithms to the
data to analyse and optimise the set of rules in use, followed by
identifying patterns in the data which could be used to write new filters
to recognise and reject spam rather than accepting it.  The parser provides
the data in a normalised form which is far easier to use as input to new or
existing implementations of machine-learning algorithms than trying to
adapt each algorithm to extract data directly from the log files.  The
current focus is on clustering and decision trees to optimise the order in
which rules are applied; future efforts will involve using data gathered by
the parser to train and test new filters.  This task is very similar to
optimising a black box application based on its inputs and outputs, and
this approach could be applied to optimising the behaviour of any system
given sufficient logging to analyse.  An alternative approach to black box
optimisation which uses application profiling in conjunction with the
application's error messages to improve the error messages shown to users
is described in~\cite{black-box-error-reporting}; profiling data may be
useful in supplementing systems which fail to provide adequate log
messages.
